---
title: "Data Manipulation with dplyrXdf"
author: "Microsoft Data Science Team"
date: "June 6, 2016"
output: html_document
---


# Introduction

## Overview | Plan

At the end of this session, you will have learned how to:

* Take advantage of the verbs and syntax you learned from the `dplyr` module to manipulate `RxXdfData` data objects
* Summarize your `RxXdfData` objects quickly and easily
* Create custom functions and use them for mutations and summarizations
* Understand where and when to use the `dplyrXdf` package and when to use functions from the `RevoScaleR` package

## Why dplyrXdf? | Simplify Your Analysis Pipeline

* The `RevoScaleR` package enables R users to manipulate data that is larger than memory
* It introduces a new data type, called an `xdf` (short for eXternal Data Frame), which are highly efficient out-of-memory objects
* However, many of the `RevoScaleR` functions have a dramatically different syntax from base R functions
* The `dplyr` package has becme exceptionally popular, due to its appealing syntax, and since it's extensible, we have created a new package called `dplyrXdf` that exposes most of the `dplyr` functionality to `xdf` objects
* Many data analysis pipelines require creating many intermediate datasets, which are only needed for their role in deriving a final dataset, but have no/little use on their own
* The `dplyrXdf` abstracts this task of file management, so that you can focus on the data itself, rather than the management of intermediate files
* Unlike `dplyr`, or other base R packages, `dplyrXdf` allows you to work with data residing _outside_ of memory, and therefore scales to datasets of arbitrary size


## Requirements | What You'll Need

* I expect that you have already covered the `dplyr` training
* Understand the *XDF* data type and how to import data to *XDF*
* If you're working on a different computer than your trianer: have (`devtools`)[github.com/hadley/devtools] (and if on a Windows machine, [Rtools](https://cran.r-project.org/bin/windows/Rtools/))

## Installing dplyrXdf

* The `dplyrXdf` package is not yet on CRAN
* You have to download it from [github](https://github.com/RevolutionAnalytics/dplyrXdf/)
  - if you're on a windows machine, install [Rtools](https://cran.r-project.org/bin/windows/Rtools/) as well
  - the `devtools` package provides a very handy function, `install_github`, for installing R packages saved in github repositories

## Create XDF from taxi data

```{r load_data, eval = TRUE}
taxi_xdf <- RxXdfData("yellow_tripdata_2015.xdf")
```

# Simplified Pipelines for Data Summaries

## Data Transforms | The rxDataStep Way

* All the functionality exposed by the `dplyrXdf` package can also be completed
by using the `rxDataStep` function in the `RevoScaleR` package included with your MRS installation
* In fact, `dplyrXdf` consists almost entirely of wrapper functions that call on other `rx` functions
* Let's compare how the workflow for adding a new column to a dataset with `rxDataStep` vs `dplyrXdf`

```{r rxdatastepway}
rxDataStep(inData = taxi_xdf,
           outFile = RxXdfData("taxi_transform.xdf"),
           transforms = list(tip_pct = tip_amount/fare_amount),
           overwrite = TRUE)


```

## Data Transforms | The rxDataStep Way

```{r rxdatastapway_continued}
rxGetInfo(RxXdfData("taxi_transform.xdf"), numRows = 2)

```

## Data Transforms | The dplyrXdf Way

* We could do the same operation with `dplyrXdf`, using the exam same syntax 
that we learned in the `dplyr` module and taking advantage of the `%>%` operator 

```{r dplyrXdf_mutate}
taxi_transform <- taxi_xdf %>% mutate(tip_pct = tip_amount/fare_amount)
taxi_transform %>% rxGetInfo(numRows = 2)

```

## Differences

* The major difference between the `rxDataStep` operation and the `dplyrXdf` method, is that we do not specify an `outFile` argument anywhere in the `dplyrXdf` pipeline
* In our case, we have assigned our `mutate` value to a new variable called `taxi_transform`
* This creates a temporary file to save the intermediate `xdf`, and only saves the most recent output of a pipeline, where a pipeline is defined as all operations starting from a raw xdf file.
* To copy an *xdf* from the temporary directory to permanent storage, use the `persist` verb

```{r inter-file}
taxi_transform@file
persist(taxi_transform, outFile = "taxiTransform.xdf")
```


## Using dplyrXdf for Aggregations | dplyrXdf Way

* The `dplyrXdf` package really shines when used for data aggregations and summarizations
* Whereas `rxSummary`, `rxCube`, and `rxCrossTabs` can compute a few summary statistics and do aggregations very quickly, they are not sufficiently general to be used in all places


```{r group-xdf}
taxi_group <- taxi_transform %>%
  group_by(pickup_nhood) %>% 
  summarize(ave_tip_pct = mean(tip_pct))
taxi_group %>% head
```

## Using dplyrXdf for Aggregations | rxCube Way

* The above could have been done with `rxCube` as well, but would require additional considerations
* We would have to make sure that the `pickup_nhood` column was a factor (can't mutate in place because of different data types)

```{r cube-way}
rxFactors(inData = taxi_transform, 
          outFile = "taxi_factor.xdf", 
          factorInfo = c("dropoff_nhood"), overwrite = TRUE)
rxCube(tip_pct ~ dropoff_nhood, means = TRUE, data = "taxi_factor.xdf")

```

