---
title: 'Introduction to R: Fundamentals, Visualization and Data Manipulation'
author: "Ali Zaidi, Machine Learning and Data Science Education Team"
date: "June 6th, 2016"
output:
  ioslides_presentation:
    logo: images/clark-logo.png
    runtime: shiny
    smaller: yes
    widescreen: yes
  html_notebook:
    toc: yes
  html_document:
    toc: yes
    keep_md: true
params: 
  prestype:
    label: "Presentation Type:"
    value: interactive
    input: select
    choices: [interactive, static]
---

# Overview of the R Project

## Lesson Plan | R U Ready?

> - What the R Language for Statistical Computing is
> - R's capabilities and it's limitations
> - What types of problems you might want to use R with
> - How to manage data with the exceptionally popular open source package dplyr
> - How to develop models and write functions in R

## What is R? | Why should I care?

* R is successor to the S Language, originated at Bell Labs AT&T
* It is based on the Scheme interpreter
* Originally designed by two University of Auckland Professors for their introductory statistics course

![Robert Gentleman and Ross Ihaka](http://revolution-computing.typepad.com/.a/6a010534b1db25970b016766fdae38970b-800wi)

## R Philosophy | What R Thou?

R follows the [Unix philosophy](http://en.wikipedia.org/wiki/Unix_philosophy)

* Write programs that do one thing and do it well. Write programs to work together.
* R is extensible with more than 9,000 packages available at CRAN (http://crantastic.org/packages)
* R, like it's inspiration, Scheme, is a functional programming language
* R evaluates lazily, allowing for a very flexible synta
* R is a highly interpreted dynamically typed language, allowing you to mutate variables and analyze datasets quickly, but is significantly slower than low-level, statically typed languages like C or Java
* R has a high memory footprint, and can easily lead to crashes if you aren't careful

## Development Environments | Where to Write R Code

* The most popular integrated development environment for R is [RStudio](https://www.rstudio.com/)
* The RStudio IDE is entirely html/javascript based, so completely cross-platform
* RStudio Server for cloud instances
* Developers of RStudio have also written a plethora of useful R packages
* For Windows machines, we have recently announced the general availability of [R Tools for Visual Studio, RTVS](https://www.visualstudio.com/en-us/features/rtvs-vs.aspx)
* RTVS will support connectivity to Azure and SQL Server in the future
* RTVS has great debugging support

# Quick Tour of Your IDE

## Strengths of R | Where R Succeeds

* Expressive
* Open source 
* Extendable -- nearly 10,000 packages with functions to use, and that list continues to grow
* Focused on statistics and machine learning -- utilized by academics and practitioners
* Advanced data structures and graphical capabilities
* Large user community, academics and industry
* It is designed by statisticians 

## Weaknesses of R | Where R Falls Short

* It is designed by statisticians
* Inefficient at element-by-element computations
* May make large demands on system resources, namely memory
* Data capacity limited by memory
* Single-threaded

## Some Essential Open Source Packages

* There are over 10,000 R packages to choose from, what do I start with?
* Data Management: `dplyr`, `tidyr`, `data.table`
* Visualization: `ggplot2`, `ggvis`, `htmlwidgets`, `shiny`
* Data Importing: `haven`, `RODBC`, `readr`, `foreign`
* Other favorites: `magrittr`, `rmarkdown`, `caret`

# R Foundations

## Command line prompts

Symbol | Meaning
------ | -------
  `>`   | ready for a new command
  `+`   | awaiting the completion of an existing command 
  
Can change these options either permanently at startup (see `?Startup`) or manually at each session with the `options` function, `options(continue = " ")` for example.

## I'm Lost! | Getting Help with R

* [Stack Overflow](http://stackoverflow.com/questions/tagged/r)
* [Cross Validated, R](http://stats.stackexchange.com/)
* [R Reference Card](https://cran.r-project.org/doc/contrib/Short-refcard.pdf)
* [RStudio Cheat Sheets](https://www.rstudio.com/resources/cheatsheets/)
* [R help mailing list and archives](https://stat.ethz.ch/mailman/listinfo/r-help)
* [CRAN Task Views](https://cran.r-project.org/web/views/)
* [Crantastic](http://crantastic.org/)
* [Revolutions Blog](http://blog.revolutionanalytics.com/)
* [RSeek](rseek.org)
* [R-Bloggers](http://www.r-bloggers.com/)

## Quick Tour of Things You Need to Know | Data Structures

* R's data structures can be described by their dimensionality, and their type.


|    | Homogeneous   | Heterogeneous |
|----|---------------|---------------|
| 1d | Atomic vector | List          |
| 2d | Matrix        | Data frame    |
| nd | Array         |               |

## Quick Tour of Things You Need to Know | Data Types

> - Atomic vectors come in one of four types
> - `logical` (boolean). Values: `TRUE` | `FALSE`
> - `integer`
> - `double` (often called numeric)
> - `character`
> - Rare types:
> - `complex` 
> - `raw`

# Lab 1: R Data Types

# Data Manipulation with the dplyr Package

## Overview

Rather than describing the nitty gritty details of writing R code, I'd like you to get started at immediately writing R code.

As most of you are data scientists/data enthusiasts, I will showcase one of the most useful data manipulation packages in R, `dplyr`.
At the end of this session, you will have learned:

* How to manipulate data quickly with `dplyr` using a very intuitive 'grammar'
* How to use `dplyr` to perform common exploratory and manipulation procedures
* How to apply your own custom functions to group manipulations `dplyr` with `mutate()`, `summarise()` and `do()`
* Connect to remote databases to work with larger than memory datasets

## Why use dplyr? | The Grammar of Data Manipulation

* R comes with a plethora of base functions for data manipulation, so why use `dplyr`?
* `dplyr` makes data manipulation easier by providing a few functions for the most common tasks and procedures
* `dplyr` achieves remarkable speed-up gains by using a C++ backend
* `dplyr` has multiple backends for working with data stored in various sources: SQLite, MySQL, bigquery, SQL Server, and many more
* `dplyr` was inspired to give data manipulation a simple, cohesive grammar (similar philosophy to `ggplot` - grammar of graphics)
* `dplyr` has inspired many new packages, which now adopt it's easy to understand syntax. 
* The recent package `dplyrXdf` brings much of the same functionality of `dplyr` to `XDF` data, and `SparkR` provides the same for manipulating Spark `DataFrames`


## Manipulation verbs

`filter`

:    select rows based on matching criteria

`slice`

:    select rows by number

`select`

:    select columns by column names

`arrange`

:    reorder rows by column values

`mutate`

:    add new variables based on transformations of existing variables

`transmute`

:    transform and drop other variables



## Aggregation verbs

`group_by`

:    identify grouping variables for calculating groupwise summary statistics


`count`

:    count the number of records per group


`summarise` | `summarize`

:    calculate one or more summary functions per group, returning one row of results per group (or one for the entire dataset)

## Viewing Data

* `dplyr` includes a wrapper called `tbl_df` makes df into a 'local df' that improves the printing of dataframes in the console (there's now a dedicated package [`tibble`](www.github.com/hadley/tibble) for this wrapper)
* if you want to see more of the data you can still coerce to `data.frame`

```{r dataimport, message = FALSE, cache = TRUE}
library(dplyr)
library(shiny)
library(stringr)
# taxi_df <- as.data.frame(data.table::fread('../data/sample_taxi.csv'))
load("../data/taxi_df.RData")
(taxi_df <- tbl_df(taxi_df))
```

# Filtering and Reordering Data

## Subsetting Data

* `dplyr` makes subsetting by rows very easy
* The `filter` verb takes conditions for filtering rows based on conditions
* **every** `dplyr` function uses a data.frame/tbl as it's first argument
* Additional conditions are passed as new arguments (no need to make an insanely complicated expression, split em up!)

```{r filter-1, cache = TRUE}
filter(taxi_df,
       dropoff_dow %in% c("Fri", "Sat", "Sun"),
       tip_amount > 1)
```

## Exercise

Your turn: 

* How many observations started in Harlem?
  - pick both sides of Harlem, including east harlem
* How many observations that started in Harlem ended in the Financial District?

## Solution

```{r harlem-fidi, echo = FALSE, eval = (params$prestype == "interactive")}
inputPanel(
  tags$textarea(id = "ex1", rows = 3, cols = 40,
                style = "width:700px;font-size:16px;",
                'filter(taxi_df)')
)
renderPrint({
  eval(parse(text = input$ex1))
})
```

## Solution

```{r harlem-fidi-sol, opts.label = 'solution'}
library(stringr)
harlem_pickups <- filter(taxi_df, str_detect(pickup_nhood, "Harlem"))
nrow(harlem_pickups)
harlem_fidi <- filter(harlem_pickups,
                      dropoff_nhood == "Financial District")
nrow(harlem_fidi)
```


## Select a set of columns

* You can use the `select()` verb to specify which columns of a dataset you want
* This is similar to the `keep` option in SAS's data step.
* Use a colon `:` to select all the columns between two variables (inclusive)
* Use `contains` to take any columns containing a certain word/phrase/character

## Select Example

```{r select1}
select(taxi_df, pickup_nhood, dropoff_nhood, fare_amount, dropoff_hour, trip_distance)
```

## Select: Other Options

starts_with(x, ignore.case = FALSE)

:    name starts with `x`

ends_with(x, ignore.case = FALSE)

:    name ends with `x`

matches(x, ignore.case = FALSE)

:    selects all variables whose name matches the regular expression `x`

num_range("V", 1:5, width = 1)

:    selects all variables (numerically) from `V1` to `V5`.

You can also use a `-` to drop variables.

## Reordering Data

* You can reorder your dataset based on conditions using the `arrange()` verb
* Use the `desc` function to sort in descending order rather than ascending order (default)

```{r arrange1}
select(arrange(taxi_df, desc(fare_amount), pickup_nhood), fare_amount, pickup_nhood)
```

## Exercise
Use `arrange()` to  sort on the basis of `tip_amount`, `dropoff_nhood`, and `pickup_dow`, with descending order for tip amount

## Solution

```{r arrange-exc, echo = FALSE, eval = (params$prestype == "interactive")}

inputPanel(
tags$textarea(id = "ex2", rows = 3, cols = 40,
              style = "width:700px;font-size:16px;",
              'arrange(taxi_df)')
)
renderPrint({
  eval(parse(text = input$ex2))
})

```

## Solution 

```{r arrange-exc-sol}
select(arrange(taxi_df, desc(tip_amount), dropoff_nhood, pickup_dow),
       tip_amount, dropoff_nhood, pickup_dow)

```


## Summary

filter

:    Extract subsets of rows. See also `slice()`

select

:    Extract subsets of columns. See also `rename()`

arrange

:    Sort your data

# Data Aggregations and Transformations

## Transformations

* The `mutate()` verb can be used to make new columns

```{r mutate}
taxi_df <- mutate(taxi_df, tip_pct = tip_amount/fare_amount)
select(taxi_df, tip_pct, fare_amount, tip_amount)
```

## Summarise Data by Groups

* The `group_by` verb creates a grouping by a categorical variable
* Functions can be placed inside `summarise` to create summary functions

```{r summarize}
summarise(group_by(taxi_df, dropoff_nhood), Num = n(), ave_tip_pct = mean(tip_pct))
```

## Group By Neighborhoods Example

```{r summarise2}
summarise(group_by(taxi_df, pickup_nhood, dropoff_nhood), Num = n(), ave_tip_pct = mean(tip_pct))
```

## Chaining/Piping

* A `dplyr` installation includes the `magrittr` package as a dependency 
* The `magrittr` package includes a pipe operator that allows you to pass the current dataset to another function
* This makes interpreting a nested sequence of operations much easier to understand

## Standard Code

* Code is executed inside-out.
* Let's arrange the above average tips in descending order, and only look at the locations that had at least 10 dropoffs and pickups.

```{r insideout}
filter(arrange(summarise(group_by(taxi_df, pickup_nhood, dropoff_nhood), Num = n(), ave_tip_pct = mean(tip_pct)), desc(ave_tip_pct)), Num >= 10)
```

## Reformatted


```{r formatinsideout}
filter(
  arrange(
    summarise(
      group_by(taxi_df, 
               pickup_nhood, dropoff_nhood), 
      Num = n(), 
      ave_tip_pct = mean(tip_pct)), 
    desc(ave_tip_pct)), 
  Num >= 10)
```

## Magrittr

* Inspired by unix `|`, and F# forward pipe `|>`, `magrittr` introduces the funny character (`%>%`, the _then_ operator)
* `%>%` pipes the object on the left hand side to the first argument of the function on the right hand side
* Every function in `dplyr` has a slot for `data.frame/tbl` as it's first argument, so this works beautifully!

```{r piping}
taxi_df %>% 
  group_by(pickup_nhood, dropoff_nhood) %>% 
  summarize(Num = n(),
            ave_tip_pct = mean(tip_pct)) %>% 
  arrange(desc(ave_tip_pct)) %>% 
  filter(Num >= 10)
```


## Pipe + group_by()

* The pipe operator is very helpful for group by summaries
* Let's calculate average tip amount, and average trip distance, controlling for dropoff day of the week and dropoff location
* First filter with the vector `manhattan_hoods`

```{r groupsum}
load("../data/manhattan.RData")
taxi_df %>% 
  filter(pickup_nhood %in% manhattan_hoods,
         dropoff_nhood %in% manhattan_hoods) %>% 
  group_by(dropoff_nhood, pickup_nhood) %>% 
  summarize(ave_tip = mean(tip_pct), 
            ave_dist = mean(trip_distance)) %>% 
  filter(ave_dist > 3, ave_tip > 0.05)

```

## Pipe and Plot

Piping is not limited to dplyr functions, can be used everywhere!

```{r pipplot}
library(ggplot2)
taxi_df %>% 
  filter(pickup_nhood %in% manhattan_hoods,
         dropoff_nhood %in% manhattan_hoods) %>% 
  group_by(dropoff_nhood, pickup_nhood) %>% 
  summarize(ave_tip = mean(tip_pct), 
            ave_dist = mean(trip_distance)) %>% 
  filter(ave_dist > 3, ave_tip > 0.05) %>% 
  ggplot(aes(x = pickup_nhood, y = dropoff_nhood)) + 
    geom_tile(aes(fill = ave_tip), colour = "white") + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = 'bottom') +
    scale_fill_gradient(low = "white", high = "steelblue")


```

## Piping to other arguments

* Although `dplyr` takes great care to make it particularly amenable to piping, other functions may not reserve the first argument to the object you are passing into it.
* You can use the special `.` placeholder to specify where the object should enter

```{r pipedot}
taxi_df %>% 
  filter(pickup_nhood %in% manhattan_hoods,
         dropoff_nhood %in% manhattan_hoods) %>% 
  group_by(dropoff_nhood, pickup_nhood) %>% 
  summarize(ave_tip = mean(tip_pct), 
            ave_dist = mean(trip_distance)) %>% 
  lm(ave_tip ~ ave_dist, data = .) -> taxi_model
summary(taxi_model)

```


## Exercise
  
Your turn: 

* Use the pipe operator to group by day of week and dropoff neighborhood
* Filter to Manhattan neighborhoods 
* Make tile plot with average fare amount in dollars as the fill

## Solution

```{r pipe-sol}

taxi_df %>%
  filter(pickup_nhood %in% manhattan_hoods,
         dropoff_nhood %in% manhattan_hoods) %>% 
  group_by(dropoff_dow, dropoff_nhood) %>% 
  summarise(ave_fare = mean(fare_amount)) %>% 
  ggplot(data = ., aes(x = dropoff_nhood, y = dropoff_dow)) + 
    geom_tile(aes(fill = ave_fare), colour = "white") + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = 'bottom') +
    scale_fill_gradient(low = "white", high = "steelblue")

```

# Functional Programming

## Creating Functional Pipelines | Reusable code

* The examples above create a rather messy pipeline operation
* The operation is pretty readable, but lacks reusability
* Since R is a functional language, we benefit by splitting these operations into functions and calling them separately
* This allows resuability; don't write the same code twice!

## Functional Pipelines | Summarization

* Let's create a function that takes an argument for the data, and applies the summarization by neighborhood to calculate average tip and trip distance

```{r taxi_summary}
taxi_hood_sum <- function(taxi_data = taxi_df, ...) {
  
  load("../data/manhattan.RData")
  taxi_data %>% 
    filter(pickup_nhood %in% manhattan_hoods,
           dropoff_nhood %in% manhattan_hoods, ...) %>% 
    group_by(dropoff_nhood, pickup_nhood) %>% 
    summarize(ave_tip = mean(tip_pct), 
              ave_dist = mean(trip_distance)) %>% 
    filter(ave_dist > 3, ave_tip > 0.05) -> sum_df
  
  return(sum_df)
  
}


```

## Functional Pipelines | Plotting Function

* We can create a second function for the plot

```{r taxiplot}
tile_plot_hood <- function(df = taxi_hood_sum()) {
  
  library(ggplot2)
  
  ggplot(data = df, aes(x = pickup_nhood, y = dropoff_nhood)) + 
    geom_tile(aes(fill = ave_tip), colour = "white") + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = 'bottom') +
    scale_fill_gradient(low = "white", high = "steelblue") -> gplot
  
  return(gplot)
}


```

## Calling Our Pipeline

* Now we can create our plot by simply calling our two functions

```{r getresults}
taxi_hood_sum(taxi_df) %>% tile_plot_hood

```

## Creating Complex Pipelines with do

* The `summarize` function is fun, can summarize many numeric/scalar quantities
* But what if you want multiple values/rows back, not just a scalar summary?
* Meet the `do` verb -- arbitrary `tbl` operations

```{r do-head}
taxi_df %>% group_by(dropoff_dow) %>% 
  arrange(desc(tip_pct)) %>% 
  do(slice(., 1:2)) %>% 
  select(dropoff_dow, tip_amount, tip_pct, 
         fare_amount, dropoff_nhood, pickup_nhood)

```

## Estimating Multiple Models with do

* A common use of `do` is to calculate many different models by a grouping variable

```{r dolm}
taxi_df %>% sample_n(10^5) %>% 
  group_by(dropoff_dow) %>% 
  do(lm_tip = lm(tip_pct ~ pickup_nhood + passenger_count + pickup_hour,
     data = .))

```

## Cleaning Output

* By design, every function in `dplyr` returns a `data.frame`
* In the example above, we get back a spooky `data.frame` with a column of `S3` `lm` objects
* You can still modify each element as you would normally, or pass it to a `mutate` function to extract intercept or statistics
* But there's also a very handy `broom` package for cleaning up such objects into `data.frames`

```{r broomlm}
library(broom)
taxi_df %>% sample_n(10^5) %>%  
  group_by(dropoff_dow) %>% 
  do(glance(lm(tip_pct ~ pickup_nhood + passenger_count + pickup_hour,
     data = .)))


```


## Summary

mutate

:    Create transformations

summarise

:    Aggregate

group_by

:    Group your dataset by levels

do

:    Evaluate complex operations on a tbl

Chaining with the `%>%` operator can result in more readable code.

## What We Didn't Cover

* There are many additional topics that fit well into the `dplyr` and functional programming landscape
* There are too many to cover in one session. Fortunately, most are well documented. The most notable omissions:
  1. Connecting to remote databases, see `vignette('databases', package = 'dplyr')`
  2. Merging and Joins, see `vignette('two-table', package = 'dplyr')`
  3. Programming with `dplyr`,`vignette('nse', package = 'dplyr')`
  4. `summarize_each` and `mutate_each`

## Thanks for Attending!

- Any questions?
- [alizaidi@microsoft.com](mailto:alizaidi@microsoft.com)